Constructing Quotient Inductive-Inductive Types

### 2018-10-30 paper.tex
some typos

A section of a displayed algebra $(N^D,z^D,s^D)$ over $(N,z,s)$ is
given by a section of the predicate $N^D$ which respects the
operations.
\begin{alignat*}{5}
  & \Delta^\S\,(N,z,s)\,(N^D,z^D,s^D) \equiv\, && (N^S:(x:N)\ra N^D\,x)\times (N^S\,z=z^D)\,\times \\
  & && \big((x:N)\ra N^D\,x\ra N^S\,(s\,x) = s^D\,x\,(N^S\,x)\big)
\end{alignat*}
The operation $\blank^\S$ is a modified dependent logical relation
interpretation: a $\Gamma$ context is interpreted as a dependent
relation $\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set$.

@@

The $\Ram$ function space is converted to metatheoretic function space
by the $\blank^\A$ operation, and $\Id$ is interpreted as metatheoretic
equality.
\[
\Phi^\A \equiv (I:\Set)\times(p:\N\ra\N\ra I)\times\big((a\,b\,c\,d:\N)\ra a+d=b+c\ra p\,a\,b=p\,c\,d\big)
\]
The initial algebra is given by the terms of type $\El\,\mathit{Int}$
in context $\Theta$. The identity type $\Id$ has the equality reflection
rule which says that if there is a term of type
$\Id\,\mathit{Int}\,t\,u$, then we have $t = u$ (conversion in the
theory of signatures is given by propositional equality in the
metatheory). Hence, terms of type $\El\,\mathit{Int}$ in $\Phi$ are
already quotiented by $\mathit{eq}$, through equality reflection.
\[
\con_\Phi \equiv \big(\Tm\,\Phi\,(\El\,\mathit{Int}),\, \lambda a\,b.\mathit{pair}\appm a\appm b,\, \lambda a\,b\,c\,d\,e.\reflect\,(\mathit{eq}\appm a\appm b\appm c\appm d\appm e)\big) : \Phi^\A
\]
The pair operator is given by the $\mathit{pair}$ variable applied
to the two natural number inputs, and the equality is given by
equality reflection on the $\mathit{eq}$ variable.

@@

\cite{altenkirch2018quotient} also concerns QIITs. There, signatures
are given by a scheme which builds up complete categories of algebras
from lists of functors. This notion of signature is more semantic than
our one, with more overhead in encoding signatures. It also does not
enforce strict positivity, hence the paper does not construct initial algebras.
Nonetheless, it is shown that well-behaved QIIT signatures
are covered by this scheme. Also, induction is shown to be equivalent
to initiality, but the notion of induction is given only up to
isomorphisms of algebras, in contrast to the current work, where it is
computed strictly.

@@

Second, the focus of the current work differs. Cartmell's main result
was the establishment of contextual categories as classifying
categories of GATs. In contrast, we do not discuss classifying
categories. Instead,
we present more constructions involving algebras; these include an
explicit initial algebra construction and the CwF model, which yields
a small internal type theory of algebras for each signature, and which
in particular allows us to exactly compute eliminators and
homomorphisms as types in the meta type theory.\footnote{This is
demoed in \cite{kaposi_et_al:LIPIcs:2018:9190}, by a Haskell program which
takes a signature as input, type checks it, then outputs the
corresponding induction methods and eliminators as an Agda file.}

@@

The motives of $\blank^{\C_\Omega}$ are given as follows (we omit the
$\Omega$ subscripts from now on).
\begin{alignat*}{5}
  & (\Gamma:\Con)^\C && : \Tms\,\Omega\,\Gamma\ra\Gamma^\A \\
  & (A:\Ty\,\Gamma)^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra\Tm\,\Omega\,(A[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra \Delta^\C\,(\sigma\circ\nu)=\sigma^\A\,(\Gamma^\C\,\nu) \\
  & (t:\Tm\,\Gamma\,A)^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra A^\C\,\nu\,(t[\nu])=t^\A\,(\Gamma^\C\,\nu)
\end{alignat*}
Once provided with a subsitution into the context, contexts are
interpreted as algebras. The interpretation of types is determined by
the need for context extension: $(\Gamma\rhd A)^\C\,(\nu,t)$ has type
$(\Gamma\rhd A)^\A\equiv(\gamma:\Gamma^\A)\times A^\A\,\gamma$. We can
provide the $\gamma$ part by $\Gamma^\C\,\nu$, so $A^\C$ has to
provide something of type $A^\A\,(\Gamma^\C\,\nu)$. Similarly, the
interpretation of substitutions and terms are determined by the
requirements of type substitution and substitution extension.

@@

When interpreting the \textbf{function space with small domain}, we
need to construct a $\Ty\,\Gamma$ from an $a : \Tm\,\Gamma\,\U$ and a
$B : \Ty\,(\Gamma\rhd\El\,a)$. The result has to internalise morphisms
from $a$ to $B$. We know that $(\Pi\,a\,B)^\A\,\gamma \equiv
(\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma, \alpha)$, which has to be the
definition of the displayed objects in the result (a displayed object
is a dependent function), since $\blank^\A$ should yield this part of
the CwF$^\K_{\Eq}$ model. Note that $(\alpha:a^\A\,\gamma)\ra
B^\A\,(\gamma, \alpha)$ is just a plain function space, without any
conditions for functiorality or structure preservation. Fortunately,
we don't need such conditions because the domain $a$ is discrete. In
the implementation, all of the required displayed CwF$^\K_{\Eq}$
structure is inherited from the codomain; for example,
$\blank\rhd\blank$ is given by using the codomain's $\blank\rhd\blank$
pointwise, and $\pi_1$ is defined by postcomposing with the codomain's
$\pi_1$. The interpretation of $\APP$ generally follows the currying
pattern seen in $\APP^\A$.

>>>>

A section of a displayed algebra $(N^D,z^D,s^D)$ over $(N,z,s)$ is
given by a section of the predicate $N^D$ which respects the
operations.
\begin{alignat*}{5}
  & \Delta^\S\,(N,z,s)\,(N^D,z^D,s^D) \equiv\, && (N^S:(x:N)\ra N^D\,x)\times (N^S\,z=z^D)\,\times \\
  & && \big((x:N)\ra N^D\,x\ra N^S\,(s\,x) = s^D\,x\,(N^S\,x)\big)
\end{alignat*}
The operation $\blank^\S$ is a modified dependent logical relation
interpretation: a context $\Gamma$ is interpreted as a dependent
relation $\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set$.

@@

The $\Ram$ function space is converted to metatheoretic function space
by the $\blank^\A$ operation, and $\Id$ is interpreted as metatheoretic
equality.
\[
\Phi^\A \equiv (I:\Set)\times(p:\N\ra\N\ra I)\times\big((a\,b\,c\,d:\N)\ra a+d=b+c\ra p\,a\,b=p\,c\,d\big)
\]
The initial algebra is given by the terms of type $\El\,\mathit{Int}$
in context $\Phi$. The identity type $\Id$ has the equality reflection
rule which says that if there is a term of type
$\Id\,\mathit{Int}\,t\,u$, then we have $t = u$ (conversion in the
theory of signatures is given by propositional equality in the
metatheory). Hence, terms of type $\El\,\mathit{Int}$ in $\Phi$ are
already quotiented by $\mathit{eq}$, through equality reflection.
\[
\con_\Phi \equiv \big(\Tm\,\Phi\,(\El\,\mathit{Int}),\, \lambda a\,b.\mathit{pair}\appm a\appm b,\, \lambda a\,b\,c\,d\,e.\reflect\,(\mathit{eq}\appm a\appm b\appm c\appm d\appm e)\big) : \Phi^\A
\]
The pair operator is given by the $\mathit{pair}$ variable applied
to the two natural number inputs, and the equality is given by
equality reflection on the $\mathit{eq}$ variable.

@@

\cite{altenkirch2018quotient} also concerns QIITs. There, signatures
are given by a scheme which builds up complete categories of algebras
from lists of functors. This notion of signature is more semantic than
ours, with more overhead in encoding signatures. It also does not
enforce strict positivity, hence the paper does not construct initial algebras.
Nonetheless, it is shown that well-behaved QIIT signatures
are covered by this scheme. Also, induction is shown to be equivalent
to initiality, but the notion of induction is given only up to
isomorphisms of algebras, in contrast to the current work, where it is
computed strictly.

@@

Second, the focus of the current work differs. Cartmell's main result
was the establishment of contextual categories as classifying
categories of GATs. In contrast, we do not discuss classifying
categories. Instead,
we present more constructions involving algebras; these include an
explicit initial algebra construction and the CwF model, which yields
a small internal type theory of algebras for each signature, and which
in particular allows us to exactly compute eliminators and
homomorphisms as types in the meta type theory.\footnote{This is
demonstrated in \cite{kaposi_et_al:LIPIcs:2018:9190}, by a Haskell program which
takes a signature as input, type checks it, then outputs the
corresponding induction methods and eliminators as an Agda file.}

@@

The motives of $\blank^{\C_\Omega}$ are given as follows (we omit the
$\Omega$ subscripts from now on).
\begin{alignat*}{5}
  & (\Gamma:\Con)^\C && : \Tms\,\Omega\,\Gamma\ra\Gamma^\A \\
  & (A:\Ty\,\Gamma)^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra\Tm\,\Omega\,(A[\nu])\ra A^\A\,(\Gamma^\C\,\nu) \\
  & (\sigma:\Tms\,\Gamma\,\Delta)^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra \Delta^\C\,(\sigma\circ\nu)=\sigma^\A\,(\Gamma^\C\,\nu) \\
  & (t:\Tm\,\Gamma\,A)^\C && : (\nu:\Tms\,\Omega\,\Gamma)\ra A^\C\,\nu\,(t[\nu])=t^\A\,(\Gamma^\C\,\nu)
\end{alignat*}
Once provided with a substitution into the context, contexts are
interpreted as algebras. The interpretation of types is determined by
the need for context extension: $(\Gamma\rhd A)^\C\,(\nu,t)$ has type
$(\Gamma\rhd A)^\A\equiv(\gamma:\Gamma^\A)\times A^\A\,\gamma$. We can
provide the $\gamma$ part by $\Gamma^\C\,\nu$, so $A^\C$ has to
provide something of type $A^\A\,(\Gamma^\C\,\nu)$. Similarly, the
interpretation of substitutions and terms are determined by the
requirements of type substitution and substitution extension.

@@

When interpreting the \textbf{function space with small domain}, we
need to construct a $\Ty\,\Gamma$ from an $a : \Tm\,\Gamma\,\U$ and a
$B : \Ty\,(\Gamma\rhd\El\,a)$. The result has to internalise morphisms
from $a$ to $B$. We know that $(\Pi\,a\,B)^\A\,\gamma \equiv
(\alpha:a^\A\,\gamma)\ra B^\A\,(\gamma, \alpha)$, which has to be the
definition of the displayed objects in the result (a displayed object
is a dependent function), since $\blank^\A$ should yield this part of
the CwF$^\K_{\Eq}$ model. Note that $(\alpha:a^\A\,\gamma)\ra
B^\A\,(\gamma, \alpha)$ is just a plain function space, without any
conditions for functorality or structure preservation. Fortunately,
we don't need such conditions because the domain $a$ is discrete. In
the implementation, all of the required displayed CwF$^\K_{\Eq}$
structure is inherited from the codomain; for example,
$\blank\rhd\blank$ is given by using the codomain's $\blank\rhd\blank$
pointwise, and $\pi_1$ is defined by postcomposing with the codomain's
$\pi_1$. The interpretation of $\APP$ generally follows the currying
pattern seen in $\APP^\A$.
